The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-10 conf_article/neon.tex
Major revision of the article, now raw but better structured.

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. Subsequently, we construct a second family of {\it F-Fanions}, which are based on the duals of convex combinations of Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that F-Muon is on par with Muon, which questions the exclusivity of a spectral norm.

@@

In this article, we tackle this question by first showing the connection between matrix norms and corresponding existing algorithms and discuss the theoretic bounds derived for those algorithms. We then leverage the family of norms dual to Ky Fan k-norms to derive a new class of algorithms with low-rank updates, which we call Fanions. Subsequently, we create a second, hybrid family named F-Fanions by constructing convex combinations of the Ky Fan norms with the Frobenius norm and taking dual of that composite norm. Working within the linear minimization oracle (LMO) framework we derive the explicit update formulas for both algorithm families. As it was done with Muon, we stipulate our algorithms to be fast to approximate, for which we utilize the Lanczos algorithm as described in \cref{sec:matrix-side}.

@@

\section{Update Step as Constrained Optimization}

\subsection{Linear Minimization Oracle Framework}

@@

In a typical neural network, the objective function $F$ depends on a set of weight matrices $\{ \mW_1, \mW_2, \ldots \}$. The optimization framework we have described is applied in a layer-wise fashion. At each iteration $k$, a stochastic gradient $g(\mX^k, \xi^k)$ is computed using a mini-batch of data with the $\xi^k$ noise via backpropagation. This yields a separate gradient component, $\mG_i^k$, for each matrix $\mX_i$. The LMO-based update rule is then applied to each matrix $\mX_i$ using its corresponding gradient component $\mG_i^k$.

@@

\subsection{Different Norms $\norm{\cdot}$ imply different updates}

In this subsection we will derive and discuss updates induced by the choice of the norm in \cref{eq:our_update}. To describe those updates, the singular value decomposition (SVD) of $\mM^k$ is required. We denote SVD of $\mM^k$ as $\mM^k = \mU \mSigma \mV^\top$, where $\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and  $\mV = [v_1, v_2, \dots, v_r]$.

\subsection{\texorpdfstring{$\normf{\mM^{k}}$ and Normalized SGD}{NSGD}}

\begin{lemma}\label{lemma:nsgd_update}
	When $\norm{\cdot} = \normf{\cdot}$, \cref{eq:our_update} turns into:

@@

\end{lemma}

The result is the same as in (Table 1, \citet{pethick2025training}), but here we use the Euclidean norm as $\normf{\mM^k} = \sqrt{\sum_{i=1}^{\min\{m,n\}}\sigma_i^2}$, which clearly is a matrix norm.

\subsection{\texorpdfstring{$\norms{\mM^{k}}$ and Muon}{Muon}}
\begin{lemma}\label{lemma:muon_update}
	When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

	\begin{equation}\label{eq:muon_update}
		\mX^{k+1} = \mX^{k} - \eta U V^\top
	\end{equation}
\end{lemma}

Though the update is well-known, we provide a much simpler proof in the appendix, when compared to \citet{bernstein2024oldoptimizernewnorm}.

@@

	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta u_1 v_1^\top
	\end{equation}
\end{lemma}
We name the derived algorithm \emph{Neon}. In the section {\it The Matrix side of updates}, we will discuss how to compute the update efficiently.

\subsection{\texorpdfstring{$\normfstar{\mM^{k}}^\dagger$ and F-Muon}{F-Muon}}
We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

@@

We name the derived algorithm \emph{F-Muon}. It turns out that F-Muon is a convex combination of Normalized SGD and Muon. The implications are significant and discussed in the following sections.

\subsection{\texorpdfstring{$\normftwo{\mM^{k}}^\dagger$ and F-Neon}{F-Neon}}
We define $\normftwo{\cdot}$ as a convex combination of $\norms{\cdot}$ and $\normf{\cdot}$:

@@

We name the derived algorithm \emph{F-Neon}. It turns out that F-Neon is a convex combination of Normalized SGD and Neon. The implications are significant and discussed in the following sections.

\subsection{\texorpdfstring{$\normkfk{\mM^k}^\dagger$ and Muon, Neon, and Centralized Dion without error feedback}{Muon, Neon, and Centralized Dion without error feedback}}
We remind the reader that the Ky Fan k-norm (\citet{bhatia2013matrix}, p. 92), which we denote as $\normkfk{\cdot}$, is the sum of the k largest singular values of the matrix. It can be proved that $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96). Special cases of the Ky Fan k-norm are the Ky Fan 1-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

	\begin{equation}\label{eq:ky_fan_update}
		\mX^{t+1} = \mX^{t} - \eta \sum_{i=1}^{k}u_i v_i^\top
	\end{equation}
\end{lemma}
In the section {\it The Matrix side of updates}, we will discuss how to compute the updates efficiently.

In addition, one can consider F-KF-k-norm: $\norm{\cdot}_{\mathrm{F-KF-k}}$ = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback. The resulting family can be named {\it F-Fanions}.

\subsection{Algorithms for Matrices $\leftrightarrow$ Algorithms for Vectors}

lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms with common $p$ may be analogous to each other, as is illustrated by \cref{tbl:mat_vs_vec_lmo}. The analogies go beyond similarities in the updates. SignSGD is very close to Adam, as noted in \citet{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well in training large models. NSGD is the same for both matrix and vector cases. Greedy Coordinate Descent methods are not applied to high-dimensional problems, from this perspective, it is not surprising that Neon underperforms on such problems.

\section{Matrix side of updates}\label{sec:matrix-side}
To compute the algorithms' updates, we use thick-restarted Lanczos method for singular value problem (TRLan) on $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ matrices (the one with less size is picked), implemented in CuPy lSibrary \citep{cupy_svds_ref} and described in \citet{simonz1998thick}.

@@

\section{Experiments}\label{sec:experiments}
\subsection{Randomized Linear Least Squares}
We first compare F-Fanions on the following L-smooth problem:

@@

@@

\appendix

\section{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
First, we need a well-known fact mentioned, for example, in \citet[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof of the fact.

Thus, we immediately find $\normfstar{\cdot}^\dagger$, which is related to F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we get
\begin{equation}\label{eq:normfstardual}
	\normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

@@

@@

\section{Updates derivations}

Proof of \cref{lemma:muon_update} follows from \cref{eq:nsgd_muon_update} with $\alpha=1$. Indeed, $\norms{\cdot}^\dagger=1\cdot\normn{\cdot} + 0 \cdot \normf{\cdot} $.

Proof of \cref{lemma:nsgd_muon_update}:
Since $\norm{\cdot}^\dagger = \normfstar{\cdot}^{\dagger\dagger} = \normfstar{\cdot}$, the goal is to reach $\normfstar{\mM^k} = \alpha \tr{\mSigma} + (1-\alpha) \normf{\mM^k}$.

>>>>

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. Subsequently, we construct a second family of {\it F-Fanions}, which are based on the duals of convex combinations of the Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that F-Muon is on par with Muon, which questions the exclusivity of a spectral norm.

@@

In this article, we tackle this question by first showing the connection between matrix norms and corresponding existing algorithms and discuss the theoretic bounds derived for those algorithms. We then leverage the family of norms dual to Ky Fan k-norms to derive a new class of algorithms with low-rank updates, which we call Fanions. Subsequently, we create a second, hybrid family named F-Fanions by constructing convex combinations of the Ky Fan norms with the Frobenius norm and taking dual of that composite norm. Working within the linear minimization oracle (LMO) framework we derive the explicit update formulas for both algorithm families. As it was done with Muon, we stipulate our algorithms to be fast to approximate, for which we utilize the Lanczos algorithm as described in ~\cref{sec:matrix-side}.

@@

\section{Preliminaries: Linear Minimization Oracle Framework}

@@

@@

In the next subsection we will derive and discuss updates induced by the choice of the norm in \cref{eq:our_update}. To describe those updates, the singular value decomposition (SVD) of $\mM^k$ is required. We denote SVD of $\mM^k$ as $\mM^k = \mU \mSigma \mV^\top$, where $\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$.

When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta U V^\top
\end{equation}.

And when $\norm{\cdot} = \normf{\cdot}$, \cref{eq:our_update} turns into:

@@

\section{Duals to Ky Fan norms instead of the spectral norm}
%\subsection{Non-operator Norms $\norm{\cdot}$ lead to new updates}

@@

	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta u_1 v_1^\top
	\end{equation}
\end{lemma}
We name the derived algorithm \emph{Neon}. In the section ~\cref{sec:matrix-side}, we will discuss how to compute the update efficiently.

\subsection{\texorpdfstring{$\normkfk{\mM^k}^\dagger$ and Muon, Neon, and Centralized Dion without error feedback}{Muon, Neon, and Centralized Dion without error feedback}}

Now: what will we get if we try to generalize?

We remind the reader that the Ky Fan k-norm (\citet{bhatia2013matrix}, p. 92), which we denote as $\normkfk{\cdot}$, is the sum of the k largest singular values of the matrix. It can be proved that $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96). Special cases of the Ky Fan k-norm are the Ky Fan 1-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

	\begin{equation}\label{eq:ky_fan_update}
		\mX^{t+1} = \mX^{t} - \eta \sum_{i=1}^{k}u_i v_i^\top
	\end{equation}
\end{lemma}
In the section ~\cref{sec:matrix-side}, we will discuss how to compute the updates efficiently.

Add more about Dion, and what its difference from us. I don't understand why their algorithm works well. I don't believe that the problem is with the error feedback.

\section{Algorithms for Matrices vs Algorithms for Vectors}
lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms with common $p$ may be analogous to each other, as is illustrated by \cref{tbl:mat_vs_vec_lmo}. The analogies go beyond similarities in the updates. SignSGD is very close to Adam, as noted in \citet{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well in training large models. NSGD is the same for both matrix and vector cases. Greedy Coordinate Descent methods are not applied to high-dimensional problems, from this perspective, it is not surprising that Neon underperforms on such problems.

\section{Conic combination of LMO-algorithms is an LMO-algorithm}

NSGD, Muon, and SignSGD work fair enough separately. Could we boost the quality if we combined them? To answer this question, we have to build a theoretical base for that.

\subsection{Duals to Convex Combination of Norm}
First, we need a well-known fact mentioned, for example, in \citet[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof of the fact.

Now we can work with different norms.

\subsection{General case}
We have norm \#1 with update \#1 and norm \#2 with update \#2. We prove the lemma that says that the update of their conic combination is a conic combination of updates \#1 and \#2.

\subsection{Frobeniusize them: F-Muon and F-Neon}
Let us consider a mixture of Fanions with another excellent LMO-algorithm, NSGD.

We define $\normfstar{\cdot}$ as a convex combination of $\normn{\cdot}$ and $\normf{\cdot}$:

@@

We name the derived algorithm \emph{F-Muon}.

We define $\normftwo{\cdot}$ as a convex combination of $\norms{\cdot}$ and $\normf{\cdot}$:

@@

We name the derived algorithm \emph{F-Neon}. F-Neon is a convex combination of Normalized SGD and Neon.

In general, one can consider F-KF-k-norm: $\norm{\cdot}_{\mathrm{F-KF-k}}$ = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback. The resulting family can be named {\it F-Fanions}.

\subsection{Add SignSGD: S-Muon and S-Neon}
Trivial formulas, but a need for more experiments.

\section{Computing the updates}\label{sec:matrix-side}
To compute the algorithms' updates, we use thick-restarted Lanczos method for singular value problem (TRLan) on $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ matrices (the one with less size is picked), implemented in CuPy lSibrary \citep{cupy_svds_ref} and described in \citet{simonz1998thick}.

@@

\section{Experiments}\label{sec:experiments}
\subsection{Randomized Linear Least Squares}
First, we compare F-Fanions on the following L-smooth problem:

@@

	{\bf Orthogonal research} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citet{riabinin2025gluon}. They could be transformed to Drop-Fanions etc. (cite Drop-Muon article), they could be viewed as Broximal Point Methods (cite the article), they could be subject to EF (cite the EF for all). Of course, they could be clipped (cite ClipScion), and they could be analyzed when the LMO is not calculated accurately (cite Beyond the Ideal).

@@

\appendix
\section{LMO for Neural Networks}
In a typical neural network, the objective function $F$ depends on a set of weight matrices $\{ \mW_1, \mW_2, \ldots \}$. The optimization framework we have described is applied in a layer-wise fashion. At each iteration $k$, a stochastic gradient $g(\mX^k, \xi^k)$ is computed using a mini-batch of data with the $\xi^k$ noise via backpropagation. This yields a separate gradient component, $\mG_i^k$, for each matrix $\mX_i$. The LMO-based update rule is then applied to each matrix $\mX_i$ using its corresponding gradient component $\mG_i^k$.

\section{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}

Based on the lemma, we immediately find $\normfstar{\cdot}^\dagger$, which is related to F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we get
\begin{equation}\label{eq:normfstardual}
	\normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

@@

Unfortunately, $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$ do not have a closed-form expression.

@@

\section{Updates derivations}
Proof of \cref{lemma:nsgd_muon_update}:
Since $\norm{\cdot}^\dagger = \normfstar{\cdot}^{\dagger\dagger} = \normfstar{\cdot}$, the goal is to reach $\normfstar{\mM^k} = \alpha \tr{\mSigma} + (1-\alpha) \normf{\mM^k}$.
